---
title: "Using binomtestGC()"
output:
  html_document:
    toc: yes
    toc_depth: 3
---

```{r include=FALSE}
require(knitr)
opts_chunk$set(tidy=FALSE,fig.width=5.5,fig.height=3.5, cache = TRUE)
```

# Preliminaries

You use `binomtestGC()` for inferential procedures regarding one population proportion $p$.  The function `proptestGC()` can also handle this analysis, but `binomtestGC()` computes exact $P$-values rather than employing the (sometimes crude) normal approximation, and it also uses a somewhat more sophisticated method to compute confidence intervals.

Make sure you have `tigerstats` loaded:

```{r message=FALSE}
require(tigerstats)
```


# Working from a Data Frame

Let

  >$p =$ the proportion of all Georgetown College students who are female
  
Suppose that you would like a 95%-confidence interval for $p$.  Since **sex** is present as a variable in the `m111survey` data frame, you can use formula-data input in the usual way to direct the attention of `binomtestGC()` to **sex** as the data.

However, when the function looks at **sex**, it sees the values "female" and "male".  It needs to be told which of them to count as a "success" when it computes the sample proportion $\hat{p}$.  You do this with the `success` argument.

So you use:

```{r}
binomtestGC(~sex,data=m111survey,
            success="female")
```

Note that you did not have to specify a confidence level:  by default, the function returns a 95%-confidence interval.

# Setting the Confidence Level

You can get intervals with other levels of confidence besides 95%, simply by setting the argument `conf.level` to the desired level (expressed as a proportion, rather than as a percentage).

For example, if you want a 90%-confidence interval for $p$, then you use:

```{r}
binomtestGC(~sex,data=m111survey,
            success="female",conf.level=0.90)
```

# Significance Tests

Let's now let

  >$p =$ the proportion of all Georgetown College students who are male

(Notice that, for variety's sake, we have switched to counting up males.)

Suppose that we want to perform a test of significance.  Then we would use:

* the argument `p` to indicate the belief of the Null Hypothesis as to the value of $p$;
* the argument `alternative` to specify the direction of the Alternative Hypothesis.  The possible values of this argument are:
    * "less"
    * "greater"
    * "two.sided" (the default value)
    
For example, if we want to test the hypotheses:

  >$H_0 : p = 0.50$
  
  >$H_a : p < 0.50$
  
then we use

```{r}
binomtestGC(~sex,data=m111survey,
            success="male",p=0.50,
            alternative="less")
```

To test the hypotheses

  >$H_0 : p = 0.50$
  
  >$H_a : p > 0.50$
  
then we use

```{r}
binomtestGC(~sex,data=m111survey,
            success="male",p=0.50,
            alternative="greater")
```

If we want a two-side test

  >$H_0 : p = 0.50$
  
  >$H_a : p \neq 0.50$


then we use

```{r}
binomtestGC(~sex,data=m111survey,
            success="male",p=0.50)
```
  
Note that there is no need to specify an alternative, since the default value of `alternative` is "two.sided".

# A Graph of the $P$-Value

We can get a graph of the $P$-value simply by setting the argument `graph` to `TRUE`:

```{r binomtesttutsexless}
binomtestGC(~sex,data=m111survey,
            success="male",p=0.50,
            alternative="less",
            graph=TRUE)
```



# Working with Summary Data

Suppose that in a poll of 2500 randomly selected registered voters, 1325 of them indicated support for the Affordable Care Act.  Suppose that we want a confidence interval for $p$, the proportion of all registered voters who favor the Act, and a two-sided test of significance with the hypotheses:

  >$H_0 : p = 0.50$
  
  >$H_a : p \neq 0.50$
  
We do not have raw data from a data frame, but the summary information we are given will suffice for `binomtestGC()`.  We need only:

* set the argument `x` to the number of successes (the count of people who said they support the Act), and
* set the argument `n` to the sample size.

Hence we use:

```{r}
binomtestGC(x=1325,n=2500,p=0.50)
```

# Want Less Output?

Sometimes you don't need to see quite so much output to the console. If you only want the basics (confidence interval for $p$ and $P$-value for your test), then set the argument `verbose` to `FALSE`.

For example, if you want a 90%-confidence interval and a two-sided test then try:

```{r}
binomtestGC(~sex,data=m111survey,
            success="male",p=0.50,
            verbose=FALSE)
```

