---
title: "Using lattice's bwplot()"
output:
  html_document:
    toc: yes
    toc_depth: 3
---

```{r include=FALSE}
require(knitr)
opts_chunk$set(tidy=FALSE,fig.width=4.5,fig.height=3.5, cache = TRUE)
```

# Preliminaries

The function `bwplot()` makes box-and-whisker plots for numerical variables.  It comes from the `lattice` package for statistical graphics, which is pre-installed with every distribution of R.  Also, package `tigerstats` depends on lattice, so if you load `tigerstats`:

```{r message=FALSE}
require(tigerstats)
```

then `lattice` will be loaded as well.


# One Numerical Variable

For a box-and-whisker plot of the fastest speeds ever driven by students in the `m111survey` data frame, use the command:

```{r bwtutfastest}
bwplot(~fastest,data=m111survey,
       xlab="speed (mph)",
       main="Fastest Speed Ever Driven")
```

Note the use of:

* the `xlab` argument to label the horizontal axis, complete with units (miles per hour);
* the `main` argument to provide a brief but descriptive title for the graph.

# Numerical and Factor Variable

Say you want to know:

  >*Who tends to drive faster:  the guys or the gals?*

Then you are studying the relationship between the numerical variable **fastest** and the factor variable **sex**.  `bwplot()` will break the fastest speeds up by sex and parallel box-and-whisker plots, if you run the following command:

```{r bwtutfastestsex}
bwplot(fastest~sex,data=m111survey,
       ylab="speed (mph)",
       main="Fastest Speed Ever Driven,\nby Sex of Subject")
```

Note the use of the "\n" to create a two-line title.  This trick can come in handy if your title is long!

If you prefer your box-and-whisker plots to be horizontal, then you can reverse the order of the variables in the formula:

```{r bwtutfastestsexhoriz}
bwplot(sex~fastest,data=m111survey,
       xlab="speed (mph)",
       main="Fastest Speed Ever Driven,\nby Sex of Subject")
```

# Warning

Box-and-Whisker plots are great for studying:

* one numerical variable
* the relationship between a numerical variable and a factor variable.

However, if you try to study the relationship between two numerical variables with `bwplot()`, you will get bizarre results.

For example, suppose you want to study the relationship between fastest speed ever driven (**fastest**) and the grade point average (**GPA**) of the subjects in `m111survey`:

```{r bwtutbad}
bwplot(fastest~GPA,data=m111survey,
       ylab="speed (mph)",
       xbal="grade-point average",
       main="Fastes Speed Ever Driven,\nby Grade-Point Average")
```

`bwplot()` expects at least one of the two variable in the formula to be numerical.  When it is presented with two numerical variables it politely makes do---apparently converting **fastest** into a new factor variable---but the resulting graph doesn't make any sense at all.

Use `xyplot()` (scatterplots) to study the relationship between two numerical variables.