---
title: "Using xtabs() and rowPerc()"
output:
  html_document:
    toc: yes
    toc_depth: 3
---

# Preliminaries

`xtabs()` is the numerical version of `barchartGC()`.  You use it when you want to study

* the distribution of one factor variable;
* the relationship between two factor variables.

The function `xtabs()` comes with the `stats` package, which is always loaded in R.  However, some of the data and other functions that we will use come from the `tigerstats` package, so make sure that it is loaded:

```{r include=FALSE}
require(tigerstats)
knitr::opts_chunk$set(cache = TRUE)
```

# One Factor Variable

To see a table of the tallies for the factor variable **seat** (from the `mat111survey` data frame in the `tigerstats` package):

```{r}
xtabs(~seat,data=m111survey)
```

In order to get the actual distribution of **seat**, you want percents rather than counts, so apply the function `rowPerc()` from `tigerstats`:

```{r}
rowPerc(xtabs(~seat,data=m111survey))
```

If you have a table of the counts for a variable, then you can enter it directly.  For example, suppose you have already made:

```{r}
Seat <- xtabs(~seat,data=m111survey)
Seat
```

Then you can just enter the table into `rowPerc()`:
```{r}
rowPerc(Seat)
```


# Two Factor Variables

You can also use `xtabs()` to study the relationship between two factor variables.  For example, if you want to see whether males and females differ in their seating preferences, then you might try formula-data input as follows:

```{r}
xtabs(~sex+seat,data=m111survey)
```

Of course, row percents are the way to actually study the relationship:

```{r}
rowPerc(xtabs(~sex+seat,data=m111survey))
```

Note the type of formula used to study the relationship between two factor variables:

$$\sim ExplanatoryFactor + ResponseFactor$$

Note also that people usually want to do several things with their two-way table, so often they will make it and store it in a well-named object, and then print out the object:

```{r}
SexSeat <- xtabs(~sex+seat,data=m111survey)
SexSeat
```

Then to get row percents they just put in the named object:

```{r}
rowPerc(SexSeat)
```


# Column Percents, Too!

We don't use column percents very often, but you can get them:

```{r}
colPerc(SexSeat)
```

