---
title: "Using pbinomGC()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Using pbinomGC()}
  %\usepackage[utf8]{inputenc}
---

```{r include=FALSE}
require(knitr)
opts_chunk$set(tidy=FALSE,fig.width=5.5,fig.height=3.5)
```


## Preliminaries

`pbinomGC()` provides a direct way to compute probabilities for binomial random variables, along with graphs for the probabilities, if you want them.  The function comes from the `tigerstats` package, so make sure that `tigerstats` is loaded:

```{r message=FALSE}
require(tigerstats)
```

## Finding P(X <= x)

Suppose that you have a binomial random variable with $n=20$ trials, and chance of success $p=0.3$ on each trial, and you would like to find the chance of getting 8 or fewer successes in your 20 trials.  In other words (letting $X$ stand for the binomial random variable), you are looking for:

$$P(X \leq 8).$$

Then run this command:

```{r}
pbinomGC(bound=8,region="below",size=20,prob=0.3)
```

Note carefully what the arguments mean, here:

* `bound=8` tells where to stop
* `region="below"` takes care of the "less than" part of the requested probability
* `size=20` tells `pbinomGC()` the number of trials for $X$
* `prob=0.3` tells `pbinomGC()` the chance of success on each trial

If you would like to see a graph of the probability distribution of $X$ along with a shaded area that represents the requested probability, then you can set the argument `graph` to `TRUE`:

```{r binomtutleq8}
pbinomGC(bound=8,region="below",
         size=20,prob=0.3,graph=TRUE)
```


## Find P(X >= x)

For the same random variable $X \sim binom(20,0.3)$, suppose we want the probability of at least 10 successes, i.e.:

$$P(X \geq 10).$$

When we want to go "above" a given number, `pbinomGC()` will *not* include that number in its computation of probabilities, so asking for

```{r eval=FALSE}
pbinomGC(bound=10,region="above",
         size=20,prob=0.3)
```

would give us the chance of $X$ being at least 11, which is not what we want.

However, since $X$ can only take on whole number values, $X \geq 10$ means exactly the same thing as $X > 9$, what we really should ask for is:

```{r binomtutgeq10}
pbinomGC(bound=9,region="above",
         size=20,prob=0.3,graph=TRUE)
```

The shaded area starts at 10, as desired.

It's a good idea to keep asking for a graph along with your output, so you can check that R is giving you exactly the probability that you need.

## Strict Inequalities:  Finding P(X < x)

For this case we also have to be careful.

For the same random variable $X \sim binom(20,0.3)$, suppose we want the probability of less than 6 successes, i.e.:

$$P(X < 6).$$

Since $X$ is a discrete random variable with whole number values, $X < 6$ means exactly the same thing thing as $X \leq 5$.  `pbinomGC()` includes the bound in the probability that it computes, so we have stop at 5, rather than at 6.  We use the command:

```{r binomtutless6}
pbinomGC(bound=5,region="below",
         size=20,prob=0.3,graph=TRUE)
```

## Strict Inequalities:  Finding P(X > x)

For the same random variable $X \sim binom(20,0.3)$, suppose we want the probability that the number of successes will be more than 7.  Then we are looking for:

$$P(X > 7).$$

This is easy again.  Since `pbinomGC()` does not include the bound when `region` is "above", we only need to ask for:


```{r binomtutgreater7}
pbinomGC(bound=7,region="above",
         size=20,prob=0.3,graph=TRUE)
```

## Finding P(a <= X <= b)

For the same random variable $X \sim binom(20,0.3)$, suppose we want the probability that the number of successes will be at least 6 but no more than 10.  Then we are looking for:

$$P(6 \leq X \leq 10).$$

We can find this with the command:

```{r binomtutbetwee}
pbinomGC(bound=c(6,10),region="between",
         size=20,prob=0.3,graph=TRUE)
```
Notice that there are no two numbers---6 and 10---that make the boundaries, so we must supply both of them to the `bound` argument as a list:  `c(6,10)`.  Also note that when `region="between"`, `pbinomGC()` includes *both* of the bounds in its computation of probabilities.

## Finding P(X = x)

To find the chance of $X$ being *exactly* a particular value, you can employ a little trick.

For example, for the same random variable $X \sim binom(20,0.3)$, suppose we want the chance of exactly 6 successes.  This is very same thing as

$$P(6 \leq X \leq 6),$$

so you can ask for:

```{r binomtutequal}
pbinomGC(bound=c(6,6),region="between",
         size=20,prob=0.3,graph=TRUE)
```



## "Outside" Probabilities

This case does not occur very often.

For the same random variable $X \sim binom(20,0.3)$, suppose we want the probability that you either get less than 7 successes OR you get more than 9 successes.  Then we are looking for:

$$P(X < 7 \textbf{ or } X > 9).$$


You can find this with the command:

```{r binomtutout}
pbinomGC(bound=c(7,9),region="outside",
         size=20,prob=0.3,graph=TRUE)
```
Notice that there are two numbers---7 and 9---that make the boundaries, so we must supply both of them to the `bound` argument as a list:  `c(6,10)`.  Notice also that when `region` is set to "outside", `pbinomGC()` excludes both bounds.
