---
title: "R:  Descriptive Statistics"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{R:  Descriptive Statistics}
  %\usepackage[utf8]{inputenc}
---

```{r include=FALSE}
require(knitr)
opts_chunk$set(tidy=FALSE,fig.width=5.5,fig.height=3.5)
```

```{r include=FALSE}
library(mosaic)
library(tigerstats)
```


# Preliminaries

This is a brief guide on how to use R and functions in `tigerstats` and related packages to do some very basic descriptive statistics.  We will give "templates" for the fucntions, accompanied by no-frills examples of their use.  Consult the function tutorials or other Help documents to learn more about the options for each function.

# One Factor Variable

## Graphics

$$barchartGC(\sim variable, data = MyData)$$

```{r }
barchartGC(~seat,data=m111survey)
```



## Numerical Summaries

`xtabs()` and `rowPerc()`:

```{r}
seating <- xtabs(~seat,data=m111survey)
seating
```

```{r}
rowPerc(seating)
```



# Two Factor Variables

## Graphics

$$barchartGC(\sim exp + resp, data = MyData)$$

```{r }
barchartGC(~sex+seat,data=m111survey)
```



## Numerical Summaries

`xtabs()` and `rowPerc()`:

```{r}
sexSeat <- xtabs(~sex+seat,data=m111survey)
sexSeat
```

```{r}
rowPerc(sexSeat)
```


# One Numeric Variable


## Graphics

`histogram()`, `densityplot()`, or `bwplot()`.

$$function(\sim variable,data=myData)$$

```{r }
densityplot(~fastest,data=m111survey)
```


## Numerical Summaries

Use `favstats()`:

```{r}
favstats(~fastest,data=m111survey)
```


# One Factor and One Numeric

## Graphics

$$histogram(\sim numeric \vert factor, data=MyData)$$

$$densityplot(\sim numeric \vert factor, data=MyData)$$

$$bwplot(numeric \sim factor, data=MyData)$$

```{r }
densityplot(~fastest|sex,data=m111survey)
```


## Numerical Summaries

`favstats()` again:

$$favstats(numeric \sim factor, data=myData)$$

```{r}
favstats(fastest~sex,data=m111survey)
```



# Two Numeric Variables

## Graphics

Scatter plots:

$$xyplot(response \sim explanatory, data = myData)$$

```{r }
xyplot(GPA~fastest,data=m111survey,type=c("p","r"))
```


## Numerical Summaries

$$lmGC(response \sim explanatory, data=myData)$$

```{r }
lmGC(GPA~fastest,data=m111survey)
```

## Prediction

```{r}
fastGPAMod <- lmGC(GPA~fastest,data=m111survey)
predict(fastGPAMod,x=100)
```





