---
title: "R:  Inferential Statistics"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{R:  Inferential Statistics}
  %\usepackage[utf8]{inputenc}
---

```{r include=FALSE}
require(knitr)
opts_chunk$set(tidy=FALSE,fig.width=5.5,fig.height=3.5)
```

```{r include=FALSE}
library(mosaic)
library(tigerstats)
```


# Preliminaries

This is a brief guide on how to use R and functions in `tigerstats` and related packages to do some very basic inferential statistics.  We will give "templates" for the functions, accompanied by no-frills examples of their use.  Consult the function tutorials or other Help documents to learn more about the options for each function.

# One Factor Variable

## Two Values

We recommend `binomtestGC()`.  The template is:

$$binomtestGC(\sim variable,data=MyData,p=NullBelief).$$

```{r eval=FALSE}
binomtestGC(~sex,data=m111survey,
            p=0.50,
            alternative="two.sided",
            success="female")
```


## More Than Two Values

`chisqtestGC()` is the function to use.  The template is:

$$chisqtestGC(\sim variable,data=MyData,p=NullProbs).$$

```{r eval=FALSE}
chisqtestGC(~seat,data=m111survey,p=c(1/3,1/3,1/3))
```


# Two Factor Variables

## Both Variables Have Two Values

If both the explanatory and response variable have two values each, then you might want `proptestGC()`.  The template is:

$$proptestGC(\sim explanaotry + response, data=MyData,p=0).$$

```{r eval=FALSE}
proptestGC(~sex+love_first,data=m111survey,
                p=0,
                alternative="two,sided")
```

## One or Both Variables Have More Than two Values

If one or more of the variables has more than two values, stick with `chisqtestGC()`.  The format is:

$$chisqtestGC(\sim explanatory + response, data=MyData).$$

```{r eval=FALSE}
chisqtestGC(~sex+seat,data=m111survey)
```

# One Numerical Variable

The function is `ttestGC()`.  Template:

$$ttestGC(\sim variable, data=MyData,mu=NullBelief).$$

```{r eval=FALSE}
ttestGC(~fastest,data=m111survey,mu=100,
          alternative="two.sided")
```

If the single numerical variable of interest is a difference of two numerical variables, then you are probably interested in the "paired" t-test:

$$ttestGC(secondVar - firstVar, data=MyData,mu=0).$$

```{r eval=FALSE}
ttestGC(height - ideal_ht,data=m111survey,
          mu=0,
          alternative="two.sided")
```


# One Numerical and One Factor Variable

We assume that the factor variable has two values.

You probably want the "2-sample t-test".  The format is:

$$ttestGC(numerical \sim factor, data=MyData,mu=0).$$

```{r eval=FALSE}
ttestGC(fastest~sex,data=m111survey,
          mu=0,
          alternative="two.sided")
```




# Two Numerical Variables

This topic is not provide for in the `tigerstats` package, but if you encounter this situation you will want to learn about `lm()`, the "grown-up" version of `lmGC()`.